---
title: "Student Scores"
author: "Group 4"
date: "9 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data


## Overarching Question
What demographic, social, and school related variables affect test scores in a high school environment? Do these factors differ in a math versus as language class?

```{r}
library(ggplot2)
d1=read.csv("student/student-mat.csv",header=TRUE)
d2=read.table("student/student-por.csv",sep=";",header=TRUE)

#uhh guys there's 386 students in both datasets so the rbind didn't exactly work
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3))
```

## Dataset Description
The dataset contains 32 Variables. Most variables are categorical across a number of factors that attempt to predict end of the semester scores.

Variable  | Description | Type | Contents 
--------------|--------------|---------
School      | Student's school | Binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira  
Sex         | Student's sex | Binary: "F" - female or "M" - male  
Age         | Student's age | Numeric: from 15 to 22  
Address     | Student's home address type | Binary: "U" - urban or "R" - rural  
Famsize     | Family size | Binary: "LE3" - less or equal to 3 or "GT3" - greater than 3  
Pstatus     | Parent's cohabitation status | Binary: "T" - living together or "A" - apart  
Medu        | Mother's education | Categorical: 0 - none,  1 - primary education, 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education  
Fedu        | Father's education | Categorical: 0 - none,  1 - primary education, 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)  
Mjob        | Mother's job | Nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other"  
Fjob        | Father's job | Nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other"  
Reason      | Reason to choose this school | Categorical: close to "home", school "reputation", "course" preference or "other")  
Guardian    | Student's guardian | Categorical: "mother", "father" or "other"  
Traveltime  | Home to school travel time | Categorical: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour  
Studytime   | Weekly study time | Categorical: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours  
Failures    | Number of past class failures | Categorical: n if 1<=n<3, else 4  
Schoolsup   | Extra educational support provided by school| Binary: yes or no  
Famsup      | Family educational support | Binary: yes or no  
Paid        | Extra paid classes within the course subject (Math or Portuguese) | Binary: yes or no  
Activities  | Extra-curricular activities | Binary: yes or no  
Nursery     | Attended nursery school as a child | Binary: yes or no  
Higher      | Students wants to attend higher education | Binary: yes or no  
Internet    | Internet access at home | Binary: yes or no  
Romantic    | Student pursuing romantic relationship | Binary: yes or no  
Famrel      | Quality of family relationships | Categorical: from 1 - very bad to 5 - excellent  
Freetime    | Free time after school | Categorical: from 1 - very low to 5 - very high  
Goout       | Frequency of going out with friends | Categorical: from 1 - very low to 5 - very high  
Dalc        | Workday alcohol consumption | Categorical: from 1 - very low to 5 - very high  
Walc        | Weekend alcohol consumption | Categorical: from 1 - very low to 5 - very high  
Health      | Current health status | Categorical : from 1 - very bad to 5 - very good  
Absences    | Number of school absences | Numeric: from 0 to 93  
G1  | First Period Grade | Numeric: from 1 to 20
G2  | Second Period Grade | Numeric: from 1 to 20
G3  | Final Grade  | Numeric: from 1 to 20

## Convert Factors to Numeric
```{r}
d2$schoolD <- d2$school
levels(d2$schoolD) <- c(0, 1)
d2$schoolD <- as.numeric(as.character(d2$schoolD))

d2$sexD <- d2$sex
levels(d2$sexD) <- c(0, 1)
d2$sexD <- as.numeric(as.character(d2$sexD))

correlations <- cor(d2[,unlist(lapply(d2, is.numeric))])
library(reshape)
melted_cormat <- melt(correlations)
melted_cormat
ggplot(melted_cormat, aes(x = X1, y = X2, fill = value)) + geom_tile()
```


## Descriptive Statistics
```{r}
#slice data in different ways
library(plyr)
round(prop.table(table(d2$sex,d2$school),margin = 2),2)

ddply(d2, .(school), summarize, mean = mean(G3), sd = sd(G3))


```



## Quick pairs plots

You'll need the package GGally for this

```{r}
nonFactors<- c(3,7,8,13,14,15,24,25,26,27,28,29,30,31,32,33) #List of nonfactor columns
pairs(d1[,c(30,31,32,33)],col=d1$Fedu)

d1.sub<-as.data.frame(d1[,c(30,31,32,33,7,8)])
#ggpairs(d1.sub)
```

## All pairwise correlations

```{r}
cor.vec<-matrix(nrow=dim(d1)[2],ncol=dim(d1)[2])
for (i in 1:dim(d1)[2]) {
  for (j in 1:dim(d1)[2]) {
    if (i != j) {
      cor.vec[i,j] <- as.numeric(try(cor(d1[,i],d1[,j]),silent=TRUE)) 
    }
  }
}

which(abs(cor.vec) > 0.4, arr.ind=TRUE)

```

##Paired t-tests

```{r}
d1$avgScore<-rowMeans(cbind(d1$G1,d1$G2,d1$G3))

d1$scoreDiff2<-d1$G3-d1$G2
d1$scoreDiff1<-d1$G2-d1$G1

d2$scoreDiff2<-d2$G3-d2$G2
d2$scoreDiff1<-d2$G2-d2$G1

t.test(d1$G1,d1$G2,paired=TRUE)
t.test(d2$G1,d2$G2,paired=TRUE)

```
Difference in score for d1(math) is insignificant, but significant for d2(portugese)

## Some quick ANOVA

Just to explore a bit further

```{r}
Edu.aov<-aov(d1$avgScore ~ d1$Fedu + d1$Medu)
Alc.aov<-aov(d1$avgScore ~ d1$Walc + d1$Dalc)
Job.aov<-aov(d1$avgScore ~ d1$Mjob + d1$Fjob)
summary(aov(d1$avgScore ~ factor(d1$studytime)))
ggplot(data=d1,aes(factor(Mjob),G1))+geom_boxplot()
plot(jitter(d1$G1),jitter(d1$G2),col=c(d1$Mjob=="health")+1)
```


## PCA

```{r}
d1.sub<-as.data.frame(d1[,c(7,8,26,27,28,30,31,32,33)])

d1.sub<-as.data.frame(d1[,nonFactors])
d1.pca<-prcomp(d1.sub,scale=TRUE,center=TRUE)

pairs(as.matrix(d1.sub)%*%as.matrix(d1.pca$rotation)[,1:7],col=d1$traveltime)
d1.pca$rotation
```

I've tried looking for some nice relationships here, nothing so far, will keep looking. Not optimistic.